---
title: "R Notebook"
output: html_notebook
---


# CodeBook - run analysis
This document explains the choices made and the steps followed to perform the analysis.

## Data
The data provided is separated in several files (e.g. activites labels).
The original measured data (in "Inertial Signals" folder) will not be used, for we will use the preprocessed data in X_train an X_test instead.

After opening the X_train and X_test files with a text editor, we see that :

- the format used is fixed width (width is 16, separator is empty string)
- the number of lines in the trainig dataset is 7352
- the number of lines in the test dataset is 2947

We know that the number of columns is the number of features, 561

We can guess that reading the data files may take some time...

We must use read.fwf function, with appropriate parameters, providing values for:

- width values (rep(16, 561))
- separator (sep="")
- number of rows to read (n=7352 or n=2947)
- column names (col.names to be retrieved from features.txt)
- column classes (rep("numeric", 561))
- buffer size (buffersize = 100 was optimal for my configuration after some tests with system.time())
Data units is standard gravity units 'g' for acc data (mean , std), radians / s for gyro data (mean, std).

After opening the subject_train, subject_test, features and activity_labels files with a text editor, we see that :

- the three first ones are one single column, the fourth uses a single space as separator
We can use read.table function for these files.

We define a function for calculating file pathes.
```{r}
fn <- function(fileName) {
        paste(".", datadir, fileName, sep = "/")
}
```

## Merge training dataset with test dataset
Merge is performed through the rbind standard function.
The resulting dataset has 10299 rows and 561 columns, and is named X.
```{r}
## Merges the training and the test sets to create one data set.
datadir <- "UCI HAR Dataset"
# Names of columns from features.txt
fnF <- fn("features.txt")
features_raw <- read.table(fnF, sep = " ", stringsAsFactors = FALSE)
features <- features_raw[, 2]

fnXtr <- fn("train/X_train.txt")
X_train <- read.fwf(fnXtr, rep(16, 561), sep="", header = FALSE, n=7352, row.names = NULL,
                   col.names = features, colClasses = rep("numeric", 561), buffersize = 100)
fnXts<- fn("test/X_test.txt")
X_test <- read.fwf(fnXts, rep(16, 561), sep="", header = FALSE, n=2947, row.names = NULL,
                   col.names = features, colClasses = rep("numeric", 561), buffersize = 100)
X <- rbind(X_train, X_test)
```


## Extract only the measurements on the mean and standard deviation for each measurement
Measurements are total acceleration and body gyro acceleration.
As far as body acceleration is calculated from gravity acceleration and total acceleration, columns to be considered must have their names :

- being prefixed with "tBodyAcc", "tGravityAcc" or "tBodyGyro"
- and containing ".mean." or ".std."

The corresponding columns are selected via a regular expression. The resulting dataset is called X_mean_std
```{r}
## Extracts only the measurements on the mean and standard deviation for each measurement
X_mean_std <- X[, grep("(.*)(tBodyAcc\\.|tGravityAcc\\.|tBodyGyro\\.)(mean|std)(.*)",
                          names(X), value = TRUE)]
```


## Use descriptive activity names to name the activities in the data set
Activity names are stored in the activity_labels.txt file.
We have to load the dataset and merge it with X dataframe.
Merge will be performed using the plyr join function.
```{r}
## Uses descriptive activity names to name the activities in the data set
fnActLbl <- fn("activity_labels.txt")
activityLabels <- read.table(fnActLbl, sep = " ", stringsAsFactors = FALSE,
                             col.names = c("numAct", "libAct"))

fnActivities_train <- fn("train/y_train.txt")
activities_train <- read.table(fnActivities_train, stringsAsFactors = FALSE, col.names = c("numAct"))
fnActivities_test <- fn("test/y_test.txt")
activities_test <- read.table(fnActivities_test, stringsAsFactors = FALSE, col.names = c("numAct"))
activities <- rbind(activities_train, activities_test)

X_mean_std_actnum <- cbind(activities, X_mean_std)

library(plyr)
mergedData <- join(activityLabels, X_mean_std_actnum)
```

## Appropriately labels the data set with descriptive variable names
Column names have to be in lower case, without any special character (no caret, no space, an so on), and describe column content
```{r}
## Appropriately labels the data set with descriptive variable names
names(mergedData) <- c("activitynumber", "activitylabel",
                       "bodyaccelerationmeanalongx", "bodyaccelerationmeanalongy", "bodyaccelerationmeanalongz",
                       "bodyaccelerationstandarddeviationalongx", "bodyaccelerationstandarddeviationalongy",
                       "bodyaccelerationstandarddeviationalongz",
                       "gravityaccelerationmeanalongx", "gravityaccelerationmeanalongy", "gravityaccelerationmeanalongz",
                       "gravityaccelerationstandarddeviationalongx", "gravityaccelerationstandarddeviationalongy",
                       "gravityaccelerationstandarddeviationalongz",
                       "bodyangularvelocitymeanalongx", "bodyangularvelocitymeanalongy",
                       "bodyangularvelocitymeanalongz",
                       "bodyangularvelocitystandarddeviationalongx", "bodyangularvelocitystandarddeviationalongy",
                       "bodyangularvelocitystandarddeviationalongz")
```


## From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject
The second dataset is obtained by using the aggregate function, it is named agg.
The results are written in a file called result_average-mean-std.txt, in the script folder.
```{r}
## From the data set in step 4, creates a second, independent tidy data set
## with the average of each variable for each activity and each subject
fnSubjects_train <- fn("train/subject_train.txt")
subjects_train <- read.table(fnSubjects_train, stringsAsFactors = FALSE,
                             col.names = c("subjectid"), colClasses = c("integer"))
fnSubjects_test <- fn("test/subject_test.txt")
subjects_test <- read.table(fnSubjects_test, stringsAsFactors = FALSE,
                            col.names = c("subjectid"), colClasses = c("integer"))
subjects <- rbind(subjects_train, subjects_test)

mm <- cbind(subjects, mergedData)

library(reshape2)

id_vars <- list(mm$subjectid, mm$activitylabel)

agg <- aggregate.data.frame(mm[, 4:21], id_vars, mean, drop = TRUE)
names(agg)[1:2] <- c("subjectid", "activitylabel")

fnResult <- fn("../result_average-mean-std.txt")
write.table(agg, fnResult, row.names = FALSE)
```


